---
title: "Missing does not equal blank: your SAS jedi mind tricks won't work here."
description: |
  I inerited a piece of R code written by a former SAS programmer. It did not work as intended.
author: Jonathan Bleier
date: 02-28-2021
output:
  distill::distill_article:
    self_contained: false
categories:
  - rstats
  - sas
  - missing data
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I came across some code at work where the author wanted to condition an action based on the presence of missing data. Many of the people I work with also started out as and still work as SAS programmers, so we have a habit of trying to do things in R like:
```
if (x == "") {do something}
```
rather than `is.na(x)`. I found the bug because the code wasn't working properly. It was supposed to do something when data was missing. I had missing data but nothing was happening as a consequence.

Let's see the difference between the the two.

```{r}
x <- c("a", "b", "", NA, "c")
y <- c(1, 2, NA, 3)
```

Find the missing data with `x == ""`

```{r}
x == ""
y == ""
```

What about `is.na()`?

```{r}
is.na(x)
is.na(y)
```
```{r}

```

So, `x == ""` is true if it is literally true. In SAS programming, we can use `""` as shorthand for missing. But with R, if a vector has a value `NA`, the logical test returns `NA`. The code I was working with failed because the missing data was `NA` so the logical test in the code did not return `TRUE`.

The function `is.na()` is the best way to find missing values in both character and numerical vectors with one small caveat; I have seen character data where missing values were recorded as `""`. If you have data like that, `is.na()` won't find it. 

